### 2016-04-22 manuscript-body.tex
Argue that a misconceptions focus is narrow

Again, the researchers are right that the student is failing to describe
the code in a way consistent with canon.\footnote{In this case, canon is
  the specifications and operations of the Java language and its
  compilers.} But, in their non-canonical thinking Student3 evidences
potentially productive insights about design. Precisely \emph{because}
there is no code stating that \textbf{turkey} and \textbf{roastBeef} and
ham are sorted into the array, the student is \emph{inferring} that to
be true. And, while that behavior is not what's happening, it
\emph{would} be sensible to design a program where specific instances of
classes \textbf{Turkey} and \textbf{Ham} and \textbf{RoastBeef} could go
into an array of type \textbf{Meat}. To do so, a designer could define
\textbf{Turkey} and \textbf{Ham} and \textbf{RoastBeef} as inheriting
from \textbf{Meat}.

Student3 has an idea about a relationship between entities where that
relationship is not specified in the code. Kaczmarczyk et al.
(Kaczmarczyk et al., 2010) focus only on the downside of the idea: the
student fails to display a proper understanding of how arrays work in
Java. Moreover, the student might be prone to similar mistakes of
inferring information that does not actually exist in code. But, there
is also an upside of this idea. Because Student3 is thinking about
real-world propositions like turkey and ham being kinds of meat, they
might be prepared to appreciate and discuss an object-oriented way to
put turkey in a \textbf{Meat} array. But, that possibility is
speculative conjecture. We can't know for certain whether Student3 could
be tipped into a productive object-oriented design activity around the
meats example because that question was not a focus of the research.

My second example of research that criticizes students' non-canonical
understandings comes from Bonar and Soloway's (1983) study of Pascal
programmers, part of which is discussed in Pea (1986). A student in
Bonar \& Soloway's study was asked to ``Write a program which reads in
ten integers and prints the average of those integers'' (Bonar \&
Soloway, 1983, p. 12). In pseudo-code, she wrote:

Repeat

(1) Read a number (Num)

(1a) Count := Count + 1

(2) Add the number to Sum

(2a) Sum := Sum + Num

(3) until Count :=10

(4) Average := Sum div Num

(5) writeln (`average = `,Average)

@@

Interviewer: Steps 1a and 2a: are those the same kinds of statements?

@@

Subject: I think of this {[}point to 1a{]} as just a constant, something
that keeps the loop under control. And this {[}points to 2a{]} has
something to do with something that you are gonna, that stores more
kinds of information that you are going to take out of the loop with
you. (Bonar \& Soloway, 1983, p. 12)

Pea's (1986) interpretation? ``Here, again, we see the student believing
that the programming language knows more about her intentions than it
possibly can'' (p. 32).

As before, this student has an idea about relationships in code. Pea
(1986) see the downside of her idea: believing PASCAL can understand
shades of programmer intent when, in fact, it cannot. And again, that
downside is real. It could cause trouble for this programmer later on if
she expects PASCAL to interpret her intent and it cannot.

In defense of the student, the question --- as asked --- is vague. Are
those statements the same \emph{to whom and in what way?} Pea (1986)
treats the data as though she meant ``the same to PASCAL.'' Indeed,
maybe she did, in which case his interpretation has traction. But,
another interpretation is that she meant to herself, or to someone else
reading the code. Those statements might not be the same \emph{to her}
because she treats (1a) as having a function of controlling iteration
while (2a)'s job is to combine two numbers into a new sum.

These two purposes, which for the sake of description I'll call
\emph{keeping control} and \emph{totaling up} are, in a sense,
different. The PASCAL compiler (and runtime) does not differentiate
them, but humans can. And, humans may well \emph{want} to differentiate
them. diSessa (1986) describes exactly this kind of differentiation as a
consequence of separating the structural understanding of a programming
language from a functional understanding of a language. As an example,
he discusses the structure/function difference with respect to
variables:

The structural aspects of a variable in a computer language are given
primarily by the rules for setting their values and for getting access
to their values. These rules apply in all contexts. In contrast, a
variable's functions might vary. Sometimes they might be described as
``a flag'' or more generally, as ``a communications device.'' At other
times a variable might function as ``a counter,'' ``data,'' or
``input.'' (diSessa, 1986, p. 202)

The student in Bonar \& Soloway's (1983) study did not show evidence of
understanding the structural similarities between (1a) and (2a) in her
pseudo-code. And, those authors as well as Pea (1986) justly insist that
similarity is important for students to understand. From a conceptual
standpoint, seeing the structural similarity constitutes a part of
``knowing'' PASCAL. But, even if knowing PASCAL were not the goal,
seeing the similarity helps one to take the perspective of a computing
agent that has no means for discerning programmer intent. Such
perspective-taking may help students avoid mistakes that arise from
over-assuming what a computer ``understands.''

The student did show evidence of understanding a functional difference
between (1a) and (2a), but Pea (1986) does not remark on that kind of
understanding at all.\footnote{Also glossed over is, to me, another
  important difference: a programmer might not know in advance which
  numbers are being passed in to the sum statement. So, in advance the
  programmer can say nothing about how the value of \textbf{Sum} will
  change as the loop iterates. In contrast, the programmer knows exactly
  how the value of Count will change with each loop iteration.} Again, I
claim this oversight is part of a subtle but observable trend in
programming misconceptions literature. While, or perhaps because
research has been so preoccupied warring with students' problematic
knowledge, it has sometimes failed to recover the productive knowledge
(or resources for building it) students have. In this example, the
student already has a grasp that syntactically similar statements could
serve different conceptual purposes. Hypothetically, the student might
use that information in design by calling up the $\Box=︎ \Box + 1$ syntactical
template when the situation seems to demand \emph{keeping control},
while calling up $\Box= ︎\Box + \mathrm{number}$ when \emph{totaling up} is the goal.
And, the idea that structurally identical symbol templates can serve
different functional and conceptual purposes fits precisely in line with
Sherin's (2001) theory of symbolic forms. For example, the symbolic
forms \emph{parts-of-a-whole} and \emph{base+change} have different
conceptual schemata. Parts-of-a-whole refers to the contributions of
component entities while base±change describes a kind of accumulation.
Specifically, the terms in base±change ``play different roles'' (Sherin,
2001, p. 534) But, the two distinct conceptual schemata share what I
would argue is the same concrete symbol template of how to write an
expression: $\Box=︎ \Box + \Box$.

The problem, for learning to program, comes in needing to fluidly
interpret and write code in languages that may demand incommensurable,
or at least distinct, conceptual schemata. As I show later in Table 4,
three current programming languages make remarkably different use of the
plus sign (+) as an operator. Crucially, some of the entailing ways to
make sense of how + works in those languages don't exist in Sherin's
(2001) catalog of conceptual schemata. In other words, I would argue
there are conceptual ways a programmer may need to think about
interpreting or writing a $\Box=︎ \Box +\Box$ symbol template that Sherin doesn't
enumerate.\footnote{One of the most obvious is the movement from seeing
  $\Box=︎ \Box +\Box$ as a statement of equality to seeing it as the storage of a
  sum to a variable.} To follow that implication, the canonical body of
knowledge about which programmers must reason is itself fractured,
because different languages design their operations around different
symbolic and conceptual metaphors.

To return to misconceptions, what drives research on students'
misconceptions is largely a need to get students to program computers
and reason about computation in ways that are canonically correct. And,
those are assuredly worthwhile goals. But, as I've argued, we can
already identify cases where a narrow misconceptions focus is silent
about or dismissive of students' useful intuitions. We can also identify
the further problems whereby unilateral emphasis on one language's canon
opposes the vocabulary of symbolic forms (and diverse conceptual
schemata) expert programmers ultimately need (Sherin, 2001) Taken in
total, that silence, dismissiveness, and narrow view of refining
knowledge perpetuates a deficit-focused discourse about student's
knowledge in computing.\footnote{Worse still, in my view, is that there
  is a pattern of researchers failing to discuss or acknowledge
  students' productive knowledge \emph{even when their own data supports
  it}.} Such a perspective also fails to address what aspects of
students reasoning might get broken by attempts to ``fix'' such
``misconceptions''.

It's important to note that misconceptions research in computing didn't
always treat students' non-canonical knowledge this way. I begin the
next section by backtracing to some of the earliest work on students'
cognitive ``bugs''. There, we find researchers talking more explicitly
about what's useful in students' non-canonical knowledge---a stance
largely absent in modern computing misconceptions research.

>>>>

Again, the researchers are right that the student is failing to describe
the code in a way consistent with canon. But, in their non-canonical thinking Student3 evidences
potentially ``\emph{productive}'' \cite{hammer_form_2002,hammer_tapping_2003} insights about design. Precisely \emph{because}
there is no code stating that \textbf{turkey} and \textbf{roastBeef} and
ham are sorted into the array, the student is \emph{inferring} that to
be true. And, while that behavior is not what's happening, it
\emph{would} be sensible to design a program where specific instances of
classes \textbf{Turkey} and \textbf{Ham} and \textbf{RoastBeef} could go
into an array of type \textbf{Meat}. To do so, a designer could define
\textbf{Turkey} and \textbf{Ham} and \textbf{RoastBeef} as inheriting
from \textbf{Meat}.

Put another way, it's true Student3 has an idea about a relationship between entities that
is not specified in the code. The study authors focus only on the \emph{downside} of the idea: the
student fails to display a proper understanding of how arrays work in
Java. But, there
is also an upside of this idea. Because Student3 is thinking about
real-world propositions like turkey and ham being kinds of meat, they
might be prepared to appreciate and discuss an object-oriented way to
put turkey in a \textbf{Meat} array. But, that possibility is
speculative conjecture. We can't know for certain whether Student3 could
be tipped into a productive object-oriented design activity around the
meats example because that question was not a focus of the research.

My second example of research that criticizes students' non-canonical
understandings comes from Bonar and Soloway's 1983 study of Pascal
programmers \cite{bonar_uncovering_1983}, data from which is also analyzed and discussed in \cite{pea_languageindependent_1986}. A student in
\cite{bonar_uncovering_1983} was asked to ``Write a program which reads in
ten integers and prints the average of those integers'' (Bonar \&
Soloway, 1983, p. 12). In pseudo-code, she wrote:

\begin{verbatim}
Repeat
(1) Read a number (Num)
    (1a) Count := Count + 1
(2) Add the number to Sum
    (2a) Sum := Sum + Num
(3) until Count :=10
(4) Average := Sum div Num
(5) writeln ('average = ',Average)
\end{verbatim}

@@

\begin{quote}
  Interviewer: Steps 1a and 2a: are those the same kinds of statements?

@@

  Subject: I think of this {[}point to 1a{]} as just a constant, something
  that keeps the loop under control. And this {[}points to 2a{]} has
  something to do with something that you are gonna, that stores more
  kinds of information that you are going to take out of the loop with
  you. \cite{bonar_uncovering_1983}
\end{quote}

Pea and Kurland's interpretation? ``Here, again, we see the student believing
that the programming language knows more about her intentions than it
possibly can'' \cite{pea_cognitive_1983}.

As in Example 1, this student has an idea about relationships in code. Pea and Kurland
\cite{pea_cognitive_1983} see the \emph{downside} of her idea: believing PASCAL can understand
shades of programmer intent when, in fact, it cannot. And again, that
downside is real. It could cause trouble for this programmer later on if
she expects PASCAL to interpret her intent and it cannot.

In defense of the student, the question --- as asked --- is vague. Are
those statements the same \emph{to whom} and \emph{in what way?} Pea and Kurland
treat the data as though she meant ``the same to PASCAL.'' Indeed,
maybe she did, in which case his interpretation has traction. But,
another interpretation is that she meant to herself or to someone else
reading the code. Those statements might not be the same \emph{to her}
because she treats (1a) as having a function of controlling iteration
while (2a)'s job is to combine two numbers into a new sum.

These two purposes, which for the sake of description I'll call
\emph{keeping control} (1a) and \emph{totaling up} (2a) are, in a sense,
different. The PASCAL compiler (and runtime) does not differentiate
them, but humans can. And, humans may well \emph{want} to differentiate
them. diSessa (1986) describes exactly this kind of differentiation as a
consequence of separating the structural understanding of a programming
language from a functional understanding of a language. As an example,
he discusses the structure/function difference with respect to
variables:

\begin{quote}
  The structural aspects of a variable in a computer language are given
  primarily by the rules for setting their values and for getting access
  to their values. These rules apply in all contexts. In contrast, a
  variable's functions might vary. Sometimes they might be described as
  ``a flag'' or more generally, as ``a communications device.'' At other
  times a variable might function as ``a counter,'' ``data,'' or
  ``input.'' \cite{disessa_models_1986}
\end{quote}

The student in \cite{bonar_uncovering_1983} did not show evidence of
understanding the structural similarities between (1a) and (2a) in her
pseudo-code. And, those authors as well as others \cite{pea_cognitive_1983} justly insist that
similarity is important for students to understand. From a conceptual
standpoint, seeing the structural similarity constitutes a part of
``knowing'' PASCAL. But, even if knowing PASCAL were not the goal,
seeing the similarity helps one to take the perspective of a computing
agent that has no means for discerning programmer intent. Such
perspective-taking may help students avoid mistakes that arise from
over-assuming what a computer ``understands.''

The student did, however, show evidence of understanding a \emph{functional} \cite{disessa_models_1986} difference
between (1a) and (2a), but Pea and Kurland do not remark on that kind of
understanding at all. Both \cite{pea_cognitive_1983,bonar_uncovering_1983} also gloss over another
important difference: a programmer might not know in advance which
numbers are being passed in to the sum statement. So, in advance the
programmer can say nothing about how the value of \textbf{Sum} will
change as the loop iterates. In contrast, the programmer knows exactly
how the value of Count will change with each loop iteration. Again, we
claim this oversight is part of a subtle but observable trend in
programming misconceptions literature.

While, or perhaps because
research has been so preoccupied warring with students' problematic
knowledge, it has sometimes failed to recover the productive knowledge
(or resources for building it) students have. In Example 2, the
student already has a grasp that syntactically similar statements could
serve different conceptual purposes. And, that understanding could, in turn, shape how they program. The student might
invoke the $\Box=︎ \Box + 1$ syntactical
template (Or what Sherin would call ``symbol template'' \cite{sherin_how_2001}) when the situation seems to demand \emph{keeping control},
while invoking $\Box= ︎\Box + \mathrm{number}$ when \emph{totaling up} is the goal.
And, the idea that structurally similar symbol templates can serve
different functional and conceptual purposes fits precisely in line with both diSessa's distinction of structural/functional understanding of a programming language \cite{disessa_models_1986} and
Sherin's theory of symbolic forms \cite{sherin_how_2001}.

An example drives home the connection between \cite{disessa_models_1986} and \cite{sherin_how_2001} as it applies to programming. The symbolic
forms \emph{parts-of-a-whole} and \emph{base+change} have different
conceptual schemata. Parts-of-a-whole refers to the contributions of
component entities while base±change describes a kind of accumulation \cite{sherin_how_2001}.
Specifically, the terms in base±change ``contribute to a whole but play different roles. One is a base value; the other is a change to that base.'' But, the two distinct conceptual schemata share what I
would argue is the same symbolic structure: $\Box=︎ \Box + \Box$ (parts of a whole) and $\Box = \Box \pm \Delta$ (base + change) \cite{sherin_how_2001}.

The problem, for learning to program, comes in needing to fluidly
interpret and write code in languages that may demand incommensurable,
or at least distinct, conceptual schemata. As I show later in Table 4,
three current programming languages make remarkably different use of the
plus sign (+) as an operator. Crucially, some of the entailing ways to
make sense of how + works in those languages don't exist in Sherin's
(2001) catalog of conceptual schemata. In other words, I would argue
there are conceptual ways a \emph{programmer} may need to think about
interpreting or writing a $\Box=︎ \Box +\Box$ symbol template that even Sherin \cite{sherin_how_2001} doesn't
enumerate.One of the most obvious, for example, is the conceptual movement from seeing
  $\Box=︎ \Box +\Box$ as a statement of equality (typical in a math class) to seeing it as the assignment of a
  sum to a variable (typical in programming). To follow that implication, the canonical body of
knowledge about which programmers must reason is itself fractured,
because different languages design their operations around different
symbolic and conceptual metaphors.

To return to misconceptions, what seems to drive research on students'
misconceptions is computing education largely a need to get students to program computers
and reason about computation in ways that are canonically correct. And,
that's a worthwhile goals. But, as I've argued, we can
already identify cases where a narrow misconceptions focus is silent
about (at best) or dismissive of (at worst) students' useful intuitions. We identified
the further problems when a unilateral emphasis on one language's canon
opposes the vocabulary of symbolic forms (and diverse conceptual
schemata) expert programmers ultimately need. \emph{Taken in
total, that silence, dismissiveness, and narrow view of refining
knowledge perpetuates a deficit-focused discourse about student's
knowledge in computing.} A knowledge-deficiency perspective also fails to address what aspects of
students reasoning might get broken by attempts to ``fix'' such
``misconceptions''.

It's important to note that misconceptions research in computing didn't
always treat students' non-canonical knowledge as a problem. I begin the
next section by backtracing to some of the earliest work on students'
cognitive ``bugs''. There, we find researchers talking more explicitly
about what's useful in students' non-canonical knowledge---a stance
largely absent in modern computing misconceptions research.
